Understanding the Causal Impact of Bikeway Installation on Urban Cycling Adoption

### 2024-12-14 paper/paper.qmd
final paper edits

---
title: "Understanding the Causal Impact of Bike Lane Installation on Urban Cycling Adoption"
subtitle: "A Difference-in-Differences Analysis Shows Protected Bike Lanes Triple Bike Share Ridership in Toronto, 2017-2023"
author: 
  - Steven Li
thanks: "Code and data are available at: [https://github.com/stevenli-uoft/Toronto_BikeShare_Causality](https://github.com/stevenli-uoft/Toronto_BikeShare_Causality)."
date: today
date-format: long
abstract: "We examine the causal effect of cycling infrastructure improvements on Bike Share Toronto ridership by analyzing how different types of bikeways influence system usage between 2017-2023. Using a difference-in-differences design with over 7.4 million rides across 1,191 bikeways, we find that infrastructure improvements led to significant increases in ridership, with post-implementation effects reaching 133.1 additional monthly rides per bikeway, translating to approximately 300,000 additional annual rides across all treated segments. Protected lanes demonstrate the strongest impact, generating 381.9 additional monthly rides - more than double the effect of on-road lanes (169.2 rides) and triple that of shared roadways (111.7 rides), while upgrades to existing infrastructure showed more immediate benefits than new installations. Our findings suggest that strategic implementation of protected infrastructure can substantially increase cycling activity, with implications for Toronto's planned addition of 100 kilometers of cycling infrastructure between 2025-2027. These results provide evidence-based guidance for cities working to encourage sustainable transportation through cycling infrastructure development."
format:
  pdf:
    toc: true
include-before: |
  \newpage
number-sections: true
bibliography: references.bib
---

@@

Our analysis indicates that infrastructure improvements significantly increase bike share ridership, with effects varying substantially by implementation approach. While we observe modest initial increases of 32.4 rides per month during the treatment year, post-implementation effects grow substantially to 133.1 additional monthly rides per bikeway (p < 0.01). In aggregate, this translates to approximately 300,000 additional annual rides across all 186 treated segments - a significant impact given that these improvements represent only 12.6% of Toronto's total bikeways. Protected bike lanes show particularly strong effects, generating 381.9 additional monthly rides post-implementation (p < 0.001) - more than twice the effect of on-road lanes (169.2 rides, p < 0.001) and triple that of shared roadways (111.7 rides, p < 0.001). Notably, these gains occurred despite treated bikeways having higher initial ridership at 209.03 monthly rides, compared to control bikeways 85.59 monthly rides pre-treatment. This suggests infrastructure improvements can substantially increase cycling activity even on popular routes.

These findings have important implications for urban transportation policy as cities work to expand sustainable mobility options. Toronto's commitment to add 100 kilometers of new cycling infrastructure between 2025 and 2027 [@cyclingNetworkPlan] represents a significant opportunity to enhance cycling adoption, particularly if implemented with attention to the differential impacts of various infrastructure types. The substantial ridership gains observed from improving just 12.6% of the network suggest that targeted infrastructure investments can have outsized impacts on system utilization. Furthermore, our finding that improvements to already-popular routes generated significant additional ridership challenges the assumption that infrastructure investments should prioritize underserved areas, highlighting the complex trade-offs cities face when allocating limited resources for cycling infrastructure.

@@

# Data {#sec-data}
## Overview
This study draws on three datasets from the City of Toronto's Open Data Portal [@torontoOpenData]: the Cycling Network (Bikeways) data [@bikeWayData], and both Bike Share Ridership and Stations data [@bikeShareData]. Together, these datasets provide a detailed view of Toronto's cycling landscape, capturing infrastructure development from 2001-2023 and ridership patterns from 2017-2023. While alternative cycling metrics exist, such as manual traffic counts and automated counter data, these sources suffer from inconsistent maintenance and ceased updates after 2019. Our selected datasets offer better temporal coverage and data quality, with regular updates ensuring current, reliable measurements of both infrastructure changes and system usage.

@@

When either change occurs, the location, infrastructure type, and year of modification are documented. A key limitation of this measurement process is its temporal granularity - while the year of each change is recorded, the specific month and day are not, constraining our ability to pinpoint exact implementation dates.

@@

### Limitations and Potential Biases
Our analysis faces two primary measurement challenges that could affect causal inference. First, the Bike Share system cannot track actual route choices between endpoints. To address this, we focus on trips originating from stations within 100 meters of bikeways, assuming proximity indicates usage. While this provides a reasonable proxy, it may introduce selection bias by excluding riders who use bikeways despite starting from more distant stations.

Second, our measure of infrastructure changes may suffer from reverse causality bias. The city might prioritize upgrades for routes that already show high cycling traffic, potentially overstating the causal effect of infrastructure improvements on usage. We attempt to mitigate this through our difference-in-differences design and by examining pre-treatment trends, but this limitation should be considered when interpreting our results.

@@

## Outcome Variables
Our primary outcome variable is seasonally-adjusted monthly ridership for each bikeway (`monthly_rides_adj`). @fig-avg-monthly-ridership shows several important patterns in ridership behavior. The raw ridership (black line) displays pronounced seasonal fluctuations, with consistent peaks during summer months and troughs in winter, reflecting Toronto's strong cycling seasonality. After applying seasonal adjustments (blue line), these cyclical patterns are smoothed while preserving the underlying growth trends. Both treatment and control groups show a general upward trend in ridership over the observation period, suggesting broader growth in cycling activity across Toronto's network.

@@

## Treatment Variables
Our main treatment variable (`treatment`) indicates whether a bikeway received an infrastructure change (either an upgrade or new installation) during our study period 2019-2021. @fig-treatment-control-count presents the distribution of bikeways between treatment and control groups, showing a larger pool of control bikeways to enable robust comparison with treated segments.

@@

To explore treatment effect heterogeneity, we examine two additional dimensions of infrastructure changes:

@@

@fig-infra-type presents distinct infrastructure patterns between groups. While protected lanes dominate both treatment and control groups, comprising 50% and 54% respectively, the distribution of other types differs notably. The treatment group shows a higher proportion of on-road lanes relative to shared roadways, suggesting a deliberate strategy of upgrading less protected infrastructure. This aligns with Toronto's broader policy goal of enhancing cyclist safety through infrastructure improvements [@toronto_council_agenda].

@fig-treatment-type-total demonstrates a nearly even split in treatment approaches, with 98 new installations and 88 upgrades among the 186 treated bikeways. This balanced distribution enables us to examine whether infrastructure expansion (new installations) versus enhancement (upgrades) generates different effects on ridership patterns. Further categorization into bikeway types, and sub-treatment groups allow us to examine whether certain types of infrastructures and types of changes yield different impacts on ridership patterns.

@tbl-summary-statistics in the Appendix provides detailed quantitative breakdowns of both our outcome variables (showing pre/post/treatment period ridership counts and adjusted means) and treatment variables (showing the distribution of protected lanes, on-road lanes, and shared roadways across treatment and control groups).

@@

# Results {#sec-results}
## Parallel Trends Analysis
A key assumption of our difference-in-differences design is that treated and control bikeways would have followed parallel trends in the absence of treatment. @fig-parallel visualizes this assumption by plotting the seasonally adjusted monthly rides for both treated and control bikeways during the pre-treatment period. While the treatment group consistently shows higher monthly average ridership levels than the control group, the trends between the two groups appear to move in parallel before treatment, particularly in the period from month -20 to month -10. We exclude the six months immediately preceding treatment from this visual analysis due to possible overlap in construction, avoiding potential anticipation effects.

@@

ggplot(parallel_plot, 
       aes(x = relative_month, 
           y = monthly_rides_adj, color = treatment)) +
  stat_summary(geom = "line", fun = mean, size = 1) +
  labs(
    x = "Relative Months (Pre-Treatment)",
    y = "Seasonally Adjusted Monthly Rides",
    color = "Treatment Status"
  )  +
  scale_color_manual(values = c("TRUE" = "#ff7f0e", "FALSE" = "#1f77b4"),
                     labels = c("Control", "Treated")) +
  theme_minimal() +
  theme(legend.position = "top",
        panel.grid.major = element_line(color = "gray85", linewidth = 0.35)
        )

@@

## Difference-in-Differences Model Estimates
@tbl-main-results presents the results from our main difference-in-differences model. The model estimates that infrastructure improvements led to a significant increase in Bike Share ridership. During the treatment year, we observe a marginally significant increase of 32.4 rides per month (p < 0.10). The effect becomes substantially larger in the post-treatment period, with an estimated increase of 133.1 rides per month (p < 0.01), compared to the pre-treatment period.

The high R-squared value of 0.785 indicates that our model explains a substantial portion of the variation in ridership. The inclusion of bikeway and time fixed effects helps control for unobserved confounders that might affect ridership patterns.

@@

modelsummary(
  list("Main DiD Estimates" = main_did),
  statistic = "s.e. = {std.error}, p = {p.value}",
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_map = c(
    "period::post:treatment" = "Post-Treatment Effect",
    "period::treatment:treatment" = "During-Treatment Effect"
  ),
)
```

@@

modelsummary(
  list("Placebo Test Estimates" = placebo_random_did),
  statistic = "s.e. = {std.error}, p = {p.value}",
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_map = c(
    "period::post:placebo_treatment" = "Post-Placebo Treatment Effect",
    "period::treatment:placebo_treatment" = "During-Placebo Treatment Effect"
  ),
)

@@

### Heterogeneous Effects
We examine treatment effect heterogeneity across different types of infrastructure and implementation strategies. @tbl-hetero-lane presents the results by bikeway type, with each coefficient representing the change in ridership compared to the pre-treatment period for that specific infrastructure type. Protected lanes show the most dramatic impact, generating an additional 381.9 rides per month post-treatment (p < 0.001) – more than twice the effect of on-road lanes (169.2 rides, p < 0.001) and more than triple the effect of shared roadways (111.7 rides, p < 0.001). The substantially larger effect of protected lanes suggests that perceived safety significantly influences riders' willingness to use Bike Share.

@@

modelsummary(
  list("Heterogenous Estimates by Bikeway Type" = did_hetero_lane),
  statistic = "s.e. = {std.error}, p = {p.value}",
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_map = c(
  "period::treatment:treatment_infra::Control" = "During-Treatment Effect: Control Bikeways",
  "period::treatment:treatment_infra::Treated-Shared" = "During-Treatment Effect: Shared Roadways",
  "period::treatment:treatment_infra::Treated-OnRoad" = "During-Treatment Effect: On-Road Lanes",
  "period::treatment:treatment_infra::Treated-Protected" = "During-Treatment Effect: Protected Lanes",
  "period::post:treatment_infra::Control" = "Post-Treatment Effect: Control Bikeways",
  "period::post:treatment_infra::Treated-Shared" = "Post-Treatment Effect: Shared Roadways",
  "period::post:treatment_infra::Treated-OnRoad" = "Post-Treatment Effect: On-Road Lanes",
  "period::post:treatment_infra::Treated-Protected" = "Post-Treatment Effect: Protected Lanes"
  )
)
```

@tbl-hetero-treat compares the effectiveness of upgrading existing infrastructure versus installing new bikeways. Upgrades to existing infrastructure generate larger and more immediate benefits, showing both a significant treatment-year effect (43.8 additional rides, p < 0.10) and a strong post-treatment effect (152.7 additional rides, p < 0.01). In contrast, newly-installed bikeways show no significant effect during the treatment year and a smaller post-treatment increase (115.5 rides, p < 0.10), suggesting that improvements to established cycling routes may be more effective at increasing ridership than creating entirely new routes.

@@

modelsummary(
  list("Heterogeneous Estimates by Implementation" = did_hetero_treat_type),
  statistic = "s.e. = {std.error}, p = {p.value}",
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_map = c(
    "period::treatment:sub_treatment_type::Upgraded" = "During-Treatment Effect: Upgraded",
    "period::treatment:sub_treatment_type::Newly-Installed" = "During-Treatment Effect: Newly Installed",
    "period::post:sub_treatment_type::Upgraded" = "Post-Treatment Effect: Upgraded",
    "period::post:sub_treatment_type::Newly-Installed" = "Post-Treatment Effect: Newly Installed"
  )
)
```

@@

## Key Empirical Findings and Their Implications
The evolution of treatment effects over time tells an important story about behavior change and infrastructure adoption. During the treatment year, we observe a modest but marginally significant increase of 32.4 rides per month (p < 0.10). This relatively small initial effect likely reflects several factors. First, infrastructure improvements often cause temporary disruption during construction, potentially suppressing immediate usage of bike share in the surrounding areas. Second, the treatment year impacts may be diluted because we only observe the year of changes, not precise implementation dates. Most importantly, the minor initial increase suggests that behavioral adaptation to new infrastructure takes time as users adjust their travel patterns and routine to new routes.

The significant increase in the post-treatment period – to 133.1 additional rides per month (p < 0.01) – shows the true potential of infrastructure improvements once fully integrated into the transportation network. This impact from pre-treatment to post-treatment periods aligns with research from the University of Colorado Denver showing that "building safe facilities for cyclists is one of the biggest factors in road safety for everyone" [@MARSHALL2019100539]. The magnitude of this change suggests that infrastructure improvements don't just make existing cyclists' trips safer, they also alter the decisions behind mode of transportation for many residents.

To put these effects in perspective, our treatment estimate of 133.1 additional monthly rides per bikeway translates to approximately 24,757 additional monthly rides across all 186 treated segments, or nearly 300,000 additional rides annually. This substantial increase is reflected in the raw data (see @tbl-summary-statistics in Appendix @sec-additional-figures) - treated segments averaged 331.92 monthly rides post-treatment compared to 209.03 pre-treatment, while control segments saw a more modest increase from 85.59 to 107.77 monthly rides over the same period. The magnitude of this effect is particularly notable given that treated segments represent only about 16% of Toronto's total bikeways (186 out of 1,191 total segments) but generated significant ridership gains, suggesting that targeted infrastructure improvements can have outsized impacts on system utilization.

@@

### Heterogenous Effect Findings
Our heterogeneous effects analysis show which interventions most effectively promote cycling. Protected lanes show remarkably strong effects, generating 381.9 additional monthly rides post-treatment (p < 0.001) – more than twice the impact of on-road lanes (169.2 rides, p < 0.001) and more than triple the effect of shared roadways (111.7 rides, p < 0.001). This substantial difference in effectiveness likely stems from:

- **Safety perceptions**: The physical separation from traffic provided by protected lanes addresses the concerns of what @nacto2016 research identifies as the "interested but concerned" majority of potential cyclists.
- **User confidence**: Protected infrastructure may encourage less experienced cyclists to try Bike Share, expanding the user base beyond confident urban cyclists.
- **Urban density patterns**: Protected lanes are more commonly implemented in denser, high-traffic areas like downtown Toronto, where higher population density and concentration of destinations naturally generate more cycling trips. This geographic distribution may partially explain their stronger treatment effects compared to other infrastructure types.

The comparison between upgrades and new installations provides equally important discoveries for implementation strategy. Upgrades to existing infrastructure generate more monthly ridership (152.7 adjusted rides), compared to newly-installing bike lanes (115.5 adjusted rides) post-treatment. Additionally, during the treatment period, upgrades to existing infrastructure show immediate effect to ridership (43.8 rides, p < 0.10), whereas new installations show no significant treatment-year effect. This pattern suggests that improving existing cycling routes may be more effective than expanding into new areas, particularly for short-term ridership gains. This finding aligns with Toronto's strategic focus during the same period as our study, on upgrading existing infrastructure to "higher-order" facilities, as outlined in the 2019-2021 Cycling Network Plan Update [@toronto_council_agenda].

@@

- **Temporal precision**: Our inability to observe exact implementation dates within treatment years likely diminishes treatment-year effects and may understate immediate post-implementation impacts.
- **Spatial relationships**: While we control for station proximity to bikeways, we cannot observe actual route choices. This may lead us to underestimate infrastructure usage if riders access lanes from more distant starting points, or even overestimate usage if riders don't use bike lanes, despite starting near one. This limitation is particularly significant in downtown areas where the density of both bike lanes and stations makes it difficult to isolate specific infrastructure effects.
- **Station placement effects**: Our results may partially reflect strategic placement of new Bike Share stations near improved infrastructure rather than purely infrastructure-driven changes in behavior. While this interaction makes it harder to isolate the direct effect of infrastructure improvements, it captures the realistic policy dynamic where bike share systems expand in tandem with cycling infrastructure.

@@

- **User surveys**: Qualitative research could help explain the mechanisms behind the stronger effects of protected lanes and how infrastructure influences mode choice decisions. Such research could particularly focus on understanding the "interested but concerned" demographic identified by @nacto2016 as vital for expanding cycling adoption.
- **Network effects**: Future studies could examine how the benefits of infrastructure improvements spill over onto connecting routes and affect system-wide ridership patterns. This could help inform Toronto's goal of creating a more connected cycling network, as outlined in the city's 2025-2027 implementation program [@cyclingNetworkPlan].
- **Neighborhood-level analysis**: Future research could examine how infrastructure effects vary across Toronto's diverse neighborhoods. Such analysis would be particularly beneficial as Toronto expands Bike Share into new areas, helping identify where different types of infrastructure improvements might have the greatest impact. This geographic heterogeneity analysis could also inform equity considerations in infrastructure deployment, ensuring benefits are distributed across the city's diverse communities.
- **Integration with public transit**: Studies could investigate how cycling infrastructure near transit hubs affects multi-modal trip patterns, particularly relevant given Toronto's efforts to integrate Bike Share with the TTC system [@ttcIntegration].

@@

### Budget Allocation
Our allocation of a \$100,000 budget, will be:

>>>>

---
title: "Understanding the Causal Impact of Bikeway Installation on Urban Cycling Adoption"
subtitle: "A Difference-in-Differences Analysis Shows Protected Bike Lanes Triple Bike Share Ridership in Toronto, 2017-2023"
author: 
  - Steven Li
thanks: "Code and data are available at: [https://github.com/stevenli-uoft/Toronto_BikeShare_Causality](https://github.com/stevenli-uoft/Toronto_BikeShare_Causality)."
date: today
date-format: long
abstract: "We investigate the causal effects of cycling infrastructure improvements on Bike Share ridership in Toronto between 2017 and 2023. Using a difference-in-differences design with over 7.4 million rides across 1,191 bikeways, we find that treated segments experience an additional 133.1 monthly ride per bikeway, translating to nearly 300,000 additional rides annually, or 40% of all rides on treated bikeways post-improvement. Protected lanes generate the largest gains, more than tripling usage compared to shared roadways, while upgrades to existing bikeways outpace the benefits of new installations.Our findings suggest that strategic implementation of protected infrastructure can substantially increase cycling activity, with implications for Toronto's planned addition of 100 kilometers of cycling infrastructure between 2025-2027. These results provide evidence-based guidance for cities working to encourage sustainable transportation through cycling infrastructure development."
format:
  pdf:
    toc: true
include-before: |
  \newpage
number-sections: true
bibliography: references.bib
---

@@

Our analysis indicates that infrastructure improvements significantly increase bike share ridership, with effects varying substantially by implementation approach. While we observe modest initial increases of 32.4 rides per month during the treatment year, post-implementation effects grow substantially to 133.1 additional monthly rides per bikeway. In aggregate, this translates to approximately 300,000 additional annual rides across all 186 treated segments, with improvements directly attributable to about 40% of the 1.48 million total rides recorded on these segments in the two years following changes. This substantial impact is particularly notable given that these improvements represent only 16% of bikeways in our study. Protected bike lanes show particularly strong effects, generating 381.9 additional monthly rides post-implementation - more than twice the effect of on-road lanes (169.2 rides) and triple that of shared roadways (111.7 rides). Notably, these gains occurred despite treated bikeways having higher initial ridership at 209.03 monthly rides, compared to control bikeways 85.59 monthly rides pre-treatment. This suggests infrastructure improvements can substantially increase cycling activity even on popular routes.

These findings have important implications for urban transportation policy as cities work to expand sustainable mobility options. Toronto's commitment to add 100 kilometers of new cycling infrastructure between 2025 and 2027 [@cyclingNetworkPlan] represents a significant opportunity to enhance cycling adoption, particularly if implemented with attention to the differential impacts of various infrastructure types. The substantial ridership gains observed from improving just 16% of the network suggest that targeted infrastructure investments can have outsized impacts on system utilization. Furthermore, our finding that improvements to already-popular routes generated significant additional ridership challenges the assumption that infrastructure investments should prioritize underserved areas, highlighting the complex trade-offs cities face when allocating limited resources for cycling infrastructure.

@@

# Data {#sec-data}
## Overview
This study draws on three datasets from the City of Toronto's Open Data Portal [@torontoOpenData]: the Cycling Network (Bikeways) data [@bikeWayData], and both Bike Share Ridership and Stations data [@bikeShareData]. Together, these datasets provide a detailed view of Toronto's cycling landscape, capturing bikeway infrastructure development from 2001-2023 and ridership patterns from 2017-2023. While alternative cycling metrics exist, such as manual and automated cycling traffic data, these sources suffer from inconsistent maintenance and ceased updates after 2019. Our selected datasets offer better temporal coverage and data quality, with regular updates ensuring current, reliable measurements of both infrastructure changes and bike share system usage.

@@

When either change occurs, the location, infrastructure type, and year of modification are documented. A key limitation of this measurement process is its temporal granularity - while the year of each change is recorded, the specific start/end date of construction are not documented, constraining our ability to pinpoint exact implementation dates.

@@

### Limitations and Potential Biases
Our analysis faces several important limitations and biases that affect both the interpretation of results and their broader applicability. First, our temporal measurement of infrastructure changes lacks precision as we observe only the year of implementation, not specific dates. This annual granularity likely understates treatment-year effects by combining pre-construction, construction, and post-completion periods within the same temporal window.

A second key limitation stems from the concurrent expansion of Toronto's Bike Share network during our study period. The system grew from 2,854 stations in 2016 to over 6,850 stations in 2020, with new stations often strategically placed near improved infrastructure [@toronto_bike_share_expansion]. While this coordinated deployment reflects real-world policy implementation, it complicates our causal interpretation, as ridership increase could be attributed to improved station accessibility, rather than bikeway infrastructure improvements. Our study design helps control for system-wide growth, but cannot fully separate these complementary effects.

Third, our study faces selection bias, as infrastructure improvements may be systematically targeted toward areas with rising cycling demand or changing demographics. Toronto's decision to upgrade specific routes likely considers factors like population density, proximity to employment centers, and existing ridership patterns. Without controlling for these neighborhood-level characteristics, our treatment effects may partially capture broader urban changes rather than purely causal infrastructure impacts.

Finally, our analysis assumes that bike share rides starting near bikeways indicate infrastructure usage. While we determine a 100-meter proximity threshold as a reasonable proxy, we cannot observe actual route choices. This limitation may lead us to under-count infrastructure usage by excluding riders who access lanes from more distant starting points, or over-count usage if riders don't ultimately use nearby lanes. This measurement challenge is especially acute in dense urban areas where multiple bikeways and stations create complex route options.

These limitations and biases suggest several promising directions for future research, particularly through user surveys that could directly measure route choices and infrastructure preferences. Additionally, more granular implementation data and neighborhood-level controls could help isolate infrastructure effects from concurrent changes in the urban environment.

@@

## Outcome Variables
Our outcome variable is seasonally-adjusted monthly ridership for each bikeway, and @fig-avg-monthly-ridership shows several important patterns in ridership behavior. The raw ridership (black line) displays pronounced seasonal fluctuations, with consistent peaks during summer months and troughs in winter, reflecting Toronto's strong cycling seasonality. These seasonal patterns remain visible even when using relative months because bikeways in our sample received treatments in different calendar years, but share the same relative month structure (e.g., -24 to +24 months around treatment). After applying seasonal adjustments (blue line), these cyclical patterns are smoothed while preserving the underlying growth trends. Both treatment and control groups show a general upward trend in ridership over the observation period, suggesting broader growth in cycling activity across Toronto's network.

@@

## Treatment Variables
Our main treatment variable indicates whether a bikeway received an infrastructure change (either an upgrade or new installation) during our study period 2019-2021. @fig-treatment-control-count presents the distribution of bikeways between treatment and control groups, showing a larger pool of control bikeways to enable robust comparison with treated segments.

@@

To explore treatment effect heterogeneity, we further examine two additional dimensions of infrastructure changes:

@@

@fig-infra-type presents distinct infrastructure patterns between groups. While protected lanes dominate both treatment and control groups, comprising 50% and 54% respectively, the distribution of other types differs notably. The treatment group shows a higher proportion of on-road lanes relative to shared roadways, suggesting a deliberate strategy of upgrading less protected infrastructure. This aligns with Toronto's broader policy goal of enhancing cyclist safety through infrastructure improvements [@toronto_council_agenda]. Further categorization into bikeway types allow us to examine whether certain infrastructure types yield different impacts on ridership patterns. @fig-treatment-type-total demonstrates a nearly even split in treatment approaches, with 98 new installations and 88 upgrades among the 186 treated bikeways. This balanced distribution enables us to examine whether newly-installed or upgraded bikeways generates different effects on ridership, allowing us to suggest targeted policy implications.

@tbl-summary-statistics in the Appendix provides detailed quantitative breakdowns of both our outcome variables (showing pre/post/treatment period ridership counts and adjusted means) and treatment variables (showing the distribution of protected lanes, on-road lanes, and shared roadways), across treatment and control groups.

@@

# Results {#sec-results}
## Parallel Trends Analysis
A key assumption of our difference-in-differences design is that treated and control bikeways would have followed parallel trends in the absence of treatment. @fig-parallel visualizes this assumption by plotting the seasonally adjusted monthly rides for both treated and control bikeways during the pre-treatment period. While the treatment group consistently shows higher monthly average ridership levels than the control group, the trends between the two groups appear to move in parallel before treatment, particularly in the period from month -20 to month -10.

@@

ggplot(parallel_plot, 
       aes(x = relative_month, 
           y = monthly_rides_adj, 
           color = treatment)) +
  stat_summary(geom = "line", fun = mean, size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 0.5, alpha = 0.5,
              linetype = "dashed") +
  labs(
    x = "Relative Months (Pre-Treatment)",
    y = "Seasonally Adjusted Monthly Rides",
    color = "Treatment Status"
  )  +
  scale_color_manual(values = c("TRUE" = "#ff7f0e", "FALSE" = "#1f77b4"),
                     labels = c("Control", "Treated")) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.major = element_line(color = "gray85", linewidth = 0.35)
  )

@@

## Difference-in-Differences Model Estimates
@tbl-main-results presents the results from our main difference-in-differences model. The model estimates that infrastructure improvements led to a significant increase in Bike Share ridership. During the treatment year, we observe a marginal increase of 32.4 rides per month (p < 0.1). The effect becomes substantially larger in the post-treatment period, with an estimated increase of 133.1 rides per month (p < 0.01), compared to the pre-treatment period.

The high R-squared adjusted value of 0.781 indicates that our model explains a substantial portion of the variation in ridership. The inclusion of bikeway and time fixed effects helps control for unobserved confounders that might affect ridership patterns.

@@

modelsummary(
  list("Main DiD Estimates" = main_did),
  statistic = "s.e. = {std.error}, p = {p.value}",
  gof_map = c("r.squared", "adj.r.squared"),
  coef_map = c(
    "period::post:treatment" = "Post-Treatment Effect",
    "period::treatment:treatment" = "During-Treatment Effect"
  )
)
```

@@

modelsummary(
  list("Placebo Test Estimates" = placebo_random_did),
  statistic = "s.e. = {std.error}, p = {p.value}",
  gof_map = c("r.squared", "adj.r.squared"),
  coef_map = c(
    "period::post:placebo_treatment" = "Post-Placebo Treatment Effect",
    "period::treatment:placebo_treatment" = "During-Placebo Treatment Effect"
  )
)

@@

### Heterogeneous Effects
We examine treatment effect heterogeneity across different types of bikeway and implementation strategies. @tbl-hetero-lane presents the results by bikeway type, with each coefficient representing the change in ridership compared to the pre-treatment period for that specific bikeway type. Protected lanes show the most impact, generating an additional 381.9 rides per month post-treatment, whereas on-road lanes and shared roadways generated an additional 169.2 and 111.7 additional monthly rides, respectively. The substantially larger effect of protected lanes suggests that perceived safety significantly influences riders' willingness to use Bike Share.

@@

modelsummary(
  list("Heterogenous Estimates by Bikeway Type" = did_hetero_lane),
  statistic = "s.e. = {std.error}, p = {p.value}",
  gof_map = c("r.squared", "adj.r.squared"),
  coef_map = c(
  "period::treatment:treatment_infra::Control" = "During-Treatment Effect: Control Bikeways",
  "period::treatment:treatment_infra::Treated-Shared" = "During-Treatment Effect: Shared Roadways",
  "period::treatment:treatment_infra::Treated-OnRoad" = "During-Treatment Effect: On-Road Lanes",
  "period::treatment:treatment_infra::Treated-Protected" = "During-Treatment Effect: Protected Lanes",
  "period::post:treatment_infra::Control" = "Post-Treatment Effect: Control Bikeways",
  "period::post:treatment_infra::Treated-Shared" = "Post-Treatment Effect: Shared Roadways",
  "period::post:treatment_infra::Treated-OnRoad" = "Post-Treatment Effect: On-Road Lanes",
  "period::post:treatment_infra::Treated-Protected" = "Post-Treatment Effect: Protected Lanes"
  )
)
```

@tbl-hetero-treat compares the effectiveness of upgrading existing infrastructure against installing new bikeways. Our results show that upgrades to existing infrastructure generate larger and more immediate benefits, showing both a significant treatment-year effect of 43.8 additional monthly rides, and a strong post-treatment effect of 152.7 additional monthly rides. In contrast, newly-installed bikeways show no significant effect during the treatment year (p = 0.280) and a smaller post-treatment increase (115.5 rides, p < 0.10), suggesting that improvements to established cycling routes may be more effective at increasing ridership than creating entirely new routes.

@@

modelsummary(
  list("Heterogeneous Estimates by Implementation" = did_hetero_treat_type),
  statistic = "s.e. = {std.error}, p = {p.value}",
  gof_map = c("r.squared", "adj.r.squared"),
  coef_map = c(
    "period::treatment:sub_treatment_type::Upgraded" = "During-Treatment Effect: Upgraded",
    "period::treatment:sub_treatment_type::Newly-Installed" = "During-Treatment Effect: Newly Installed",
    "period::post:sub_treatment_type::Upgraded" = "Post-Treatment Effect: Upgraded",
    "period::post:sub_treatment_type::Newly-Installed" = "Post-Treatment Effect: Newly Installed"
  )
)
```

@@

## Key Empirical Findings and Their Implications
The evolution of treatment effects over time tells an important story about behavior change and infrastructure adoption. During the treatment year, we observe a modest increase of 32.4 rides per month, likely reflecting several factors. First, infrastructure improvements often cause temporary disruption during construction, potentially suppressing immediate usage of bike share in the surrounding areas. Second, the treatment year impacts may be diluted because we only observe the year of changes, not precise implementation dates. Most importantly, the minor initial increase suggests that behavioral adaptation to new infrastructure takes time as users adjust their travel patterns and routine to new routes.

The significant increase in the post-treatment period to 133.1 additional rides per month shows the true potential of infrastructure improvements once fully integrated into the transportation network. This impact from pre-treatment to post-treatment periods aligns with research from the University of Colorado Denver showing that "building safe facilities for cyclists is one of the biggest factors in road safety for everyone" [@MARSHALL2019100539]. The magnitude of this change suggests that infrastructure improvements don't just make existing cyclists' trips safer, they also alter the decisions behind mode of transportation for many residents.

To put these effects in perspective, our treatment estimate of 133.1 additional monthly rides per bikeway translates to approximately 24,757 additional monthly rides across all 186 treated segments, or nearly 300,000 additional rides annually. Given that treated bikeways recorded about 1.48 million total rides in the two years following improvements, our estimates suggest that 40% of these rides can be linked directly to the new or upgraded infrastructure. This substantial increase is reflected in the raw data (see @tbl-summary-statistics in Appendix @sec-additional-figures), where treated segments averaged 331.92 monthly rides post-treatment compared to 209.03 pre-treatment, while control segments saw a more modest increase from 85.59 to 107.77 monthly rides over the same period. The magnitude of this effect is particularly notable given that treated segments represent only about 16% of total bikeways in our study (186 out of 1,191 total segments) but generated significant ridership gains, suggesting that targeted infrastructure improvements can have outsized impacts on system utilization.

@@

### Heterogenous Effect Findings
Our heterogeneous effects analysis show which interventions most effectively promote cycling. Protected lanes show remarkably strong effects, generating 381.9 additional monthly rides post-treatment, more than twice the impact of on-road lanes, and more than triple the effect of shared roadways. This substantial difference in effectiveness likely stems from:

- **Safety perceptions**: The physical separation from traffic provided by protected lanes addresses the concerns of what @nacto2016 research identifies as the "interested but concerned" majority of potential cyclists.
- **User confidence**: Protected infrastructure may encourage less experienced cyclists to try Bike Share, expanding the user base beyond confident urban cyclists.
- **Urban density patterns**: Protected lanes are more commonly implemented in denser high-traffic areas like downtown Toronto, where higher population density and concentration of destinations naturally generate more cycling trips. This geographic distribution may partially explain their stronger treatment effects compared to other infrastructure types.

The comparison between upgrades and new installations provides equally important discoveries for implementation strategy. Upgrades to existing infrastructure generate more monthly ridership (152.7 rides), compared to newly-installed bikeways (115.5 rides) post-treatment. Additionally, during the treatment period, upgrades to existing infrastructure show immediate effect to ridership, whereas new installations show no significant treatment-year effect. This pattern suggests that improving existing cycling routes may be more effective than expanding into new areas, particularly for short-term ridership gains. This finding aligns with Toronto's strategic focus during the same period as our study, on upgrading existing infrastructure to "higher-order" facilities, as outlined in the 2019-2021 Cycling Network Plan Update [@toronto_council_agenda].

@@

- **Temporal precision**: Our inability to observe exact implementation dates within treatment years likely diminishes treatment-year effects and may understate immediate post-implementation impacts.
- **Spatial relationships**: While we control for station proximity to bikeways, we cannot observe actual route choices. This may lead us to underestimate infrastructure usage if riders access bikeways from more distant starting points, or even overestimate usage if riders don't use bikeways, despite starting near one. This limitation is particularly significant in downtown areas where the density of both bikeways and stations makes it difficult to isolate specific infrastructure effects.
- **Station placement effects**: Our results may partially reflect strategic placement of new Bike Share stations near improved infrastructure rather than purely infrastructure-driven changes in behavior. While this interaction makes it harder to isolate the direct effect of infrastructure improvements, it captures the realistic policy dynamic where bike share systems expand in tandem with cycling infrastructure.

@@

- **User surveys**: Qualitative research could help explain the mechanisms behind the stronger effects of protected lanes and how infrastructure influences mode choice decisions. Such research could particularly focus on understanding the "interested but concerned" demographic identified by @nacto2016 as vital for expanding cycling adoption.
- **Network effects**: Future studies could examine how the benefits of infrastructure improvements spill over onto connecting routes and affect system-wide ridership patterns. This could help inform Toronto's goal of creating a more connected cycling network, as outlined in the city's 2025-2027 implementation program [@cyclingNetworkPlan].
- **Neighborhood-level analysis**: Future research could examine how infrastructure effects vary across Toronto's diverse neighborhoods. Such analysis would address our studies selection bias and be particularly beneficial as Toronto expands Bike Share into new areas, helping identify where different types of infrastructure improvements might have the greatest impact. This geographic heterogeneity analysis could also inform equity considerations in infrastructure deployment, ensuring benefits are distributed across the city's diverse communities.
- **Integration with public transit**: Studies could investigate how cycling infrastructure near transit hubs affects multi-modal trip patterns, particularly relevant given Toronto's efforts to integrate Bike Share with the TTC system [@ttcIntegration].

@@

### Budget Allocation
With an allocation of a \$100,000 budget, and to ensure we reach our target sample size, we will distribute our budget as follows:
